---
title: "my_notes_osm"
author: "Remek"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(osmdata)
library(osmextract)
library(sf)
library(mapview)
```

<https://ogh23.robinlovelace.net/osm>

OSM is Wikipedia for maps.
It focuses on human-made infrastructure.
It's not even across space, such as there is a stron community in Berlin, but not that stron in North Korea...

<https://nptscot.github.io> uses OSM data in production on a big scale.
It concerns cycling potential.
Red colour on the map show potential bootlenecks - where there are difficulties in cycling but high need for cycling.

<https://a-b-street.github.io/osm2streets> <https://a-b-street.github.io/docs/software/abstreet.html> <https://play.abstreet.org/> use case - city simulation, all on OSM data

# Getting OSM data

<https://overpass-turbo.eu> is a OSM query tool

## We can get OSM data from Geofabrik

These are zones that OSM uses to devide the world.
They have levels like 1, 2, 3 etc.

```{r}
par(mar = rep(0, 4))

plot(osmextract::geofabrik_zones[geofabrik_zones$level == 1, "name"],
     key.pos = NULL, main = NULL)
```

```{r}
plot(geofabrik_zones[geofabrik_zones$level == 2, "name"],
     key.pos = NULL, main = NULL)
```

```{r}
plot(geofabrik_zones[geofabrik_zones$level == 3, "name"],
     key.pos = NULL, main = NULL)
```

Other way to get such zones to use \_fr pkg.
Why to use such zones from pkgs?
Because there are API limits, that may be meet when you use one big city.

```{r}
plot(openstreetmap_fr_zones[openstreetmap_fr_zones$level == 1, "name"], key.pos = NULL, main = NULL)
```

```{r}
ids_2 = openstreetmap_fr_zones$parent %in% "france"
ids_3 = openstreetmap_fr_zones$parent %in% openstreetmap_fr_zones$id[ids_2]

plot(openstreetmap_fr_zones[ids_2 | ids_3, "name"], key.pos = NULL, main = NULL)
```

bbbike provider is based on <https://download.bbbike.org/osm/bbbike/>.
It is quite different from any other provider supported in osmextract since it contains OSM data for more than 200 cities worldwide.
**bbbike provider is the safest choice if you are looking for OSM data relative to a particular city in the world.**

```{r}
osmextract::oe_match("Poznań", provider = "bbbike")
```

```{r}
osmextract::oe_match("Lodz", provider = "bbbike")

```

# download

```{r}
poznan <- osmextract::oe_get("Poznan", provider = "bbbike", 
                             return_gpkg = FALSE)
```

For the rest of exercises we will use Monaco

```{r}
monaco_osm_points = oe_get("monaco", provider = "bbbike", layer = "points")
```

```{r}
monaco_osm_lines = oe_get("monaco", provider = "bbbike", layer = "lines")
monaco_osm_mlines = oe_get("monaco", provider = "bbbike", layer = "multilinestrings")
monaco_osm_polygons = oe_get("monaco", provider = "bbbike", layer = "multipolygons")
monaco_osm_other = oe_get("monaco", provider = "bbbike", layer = "other_relations")
```

Check downloaded files

```{r}
f = list.files(oe_download_directory(), pattern = "monaco", full.names = TRUE)
f
monaco = sf::read_sf(f[1])
monaco = sf::read_sf(f[2], layer = "lines")
```

Remember that the same entity may exist as a point and as a polygon, like supermarket.

People like mapping roads, so OSM has very well road covergae.

There is no elevation in OSM.

OSM has points, lines and relations.
After importing with GDAL we get points, lines, multilinestrings, multipoligons and other_relations.
We specify these in import functions.

```{r}
plot(monaco_osm_lines)
```

```{r}
mapview(monaco_osm_lines)
```

```{r}
mapview(monaco_osm_mlines)
```

```{r}
sf::st_bbox(monaco_osm_mlines)
```

```{r}
st_bbox(monaco_osm_lines)
```

```{r}
ggplot() + 
  geom_sf(data = select(monaco_osm_lines, highway))
```

Walkabilty / cyclability <https://github.com/acteng/cyclability/> - interesting, maybe **we have illegal trash dumps where the walkability is low?**

This is an example of 10m resolution buildings raster for EU: <https://stac.ecodatacube.eu/lcv_building.residential_osm/collection.json>

<https://docs.ohsome.org/ohsome-api/stable/response-parameters.html> OSM completeness assesment tools, paper <https://www.nature.com/articles/s41467-023-39698-6>
