---
title: "notes tidy session"
author: "Remek"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Initial notes

The term *geocomputation* was coined at the university of Leeds

www.npt.scot online app for bike traffic modelling (where to build bike lanes?)

Author with Jakub Nowosad <https://r.geocompx.org/>

```{r}
pkgs = c(
  "sf",
  "tidyverse",
  "geos",
  "ggspatial",
  "spData"
)

sapply(pkgs, require, character.only = TRUE)

```

Check spData

```{r}
library(geos)
library(tidyverse)
library(sf)
library(spData)
library(ggspatial)

spData::world %>% names()
```

Print world object info

```{r}
world
```

We have WGS 84.

# Get data for Poland and neighborhoods

```{r}
poland <- world %>% 
  filter(name_long == "Poland") %>% 
  select(name_long, area_km2, pop)
```

How to remove geometry column?
It's sticky.

```{r}
class(poland)

sf::st_drop_geometry(poland) %>% class() # no sf 
```

```{r}
plot(poland, max.plot = 1)
```

*Ask questions on R Spatial Mattermost Channel*

Read about centroids - <https://en.wikipedia.org/wiki/Geographical_centre>

Make centroids and select countries:

```{r}
?st_centroid
```

Write my own code from examples in tidy.qdm

\#####################################

# GEOS

GEOS is a backend for sf that does

```{r}

```

Talks about benchmarks - they doesn't really matter as long as we are under few minutes.
For very large computations use 1% sample and benchmark.
Computation time is getting cheap, while code writing time is quite expensive, this is something that Hadley says to legitimize tidyverse.
Tidyverse is slow to run but fast and easy to write.
Geos is much faster than sf.

# Poznań data example

I did downloading yesterday.

```{r}
list.files("data/osm")
```

Read data in

```{r}
pol_all <- read_sf("./data/osm/gis_osm_transport_a_free_1.shp")

glimpse(pol_all)
```

Plot

```{r}
plot(pol_all, max.plot = 1)
```

ggplot it

```{r}
pol_all %>% ggplot() + geom_sf()
```
Filter

```{r}
pol_bus <- pol_all %>% 
  filter(str_detect(name, "bus"))


plot(pol_bus)
```
Mapview interactive 

```{r}
library(mapview)

mapview::mapview(pol_all)


```

Nice, do it on my data from e5. 

# Buffers TODO

First load text data and convert it to sf 
```{r}
stops_raw <- read_csv("./data/gtfs/stops.txt")
stops <- stops_raw %>% 
  select(-stop_code) %>% 
  st_as_sf(., coords = c("stop_lon", "stop_lat"))

st_crs(stops) # check coord ref system - none from txt
st_crs(stops) = "EPSG:4326" # set common crs

sf::sf_use_s2() # check whether is it working


```

Now do buffers 

```{r}
poi_buffers <- st_buffer(poi_sf, 150)
```

Plot buffers
```{r}
mapview(poi_buffers)
```

Now subset bus stops in these buffers: 
```{r}
stops_in_buffer <- stops[poi_buffers, ] # base R!
dim(stops_in_buffer)
```
Plot? 
```{r}
ggplot() +
  geom_sf(data = stops_in_buffer) + 
  geom_sf(data = poi_buffers, 
          fill = "darkgreen", alpha = .4)

```


# OSM and data frames 

```{r}

poi_df = tribble(
  ~name, ~lon, ~lat,
  "Faculty",        16.9418, 52.4643,
  "Hotel ForZa",    16.9474, 52.4436,
  "Hotel Lechicka", 16.9308, 52.4437,
  "FairPlayce",     16.9497, 52.4604
)
poi_sf = sf::st_as_sf(poi_df, coords = c("lon", "lat"))
sf::st_crs(poi_sf) = "EPSG:4326"
```

```{r}
mapview(poi_sf)
```

# Spatial join 

```{r}
st_join(poi_buffers, stops)
```

# Raster data 

Time flies when you have fun... 

